Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-19 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Throughout this proposal, we also refer to chemically-driven erosion as dissolution. Some examples in the figure include the rough surface of limestone pavement, a sinkhole, an erosion-curved conduit, and finally an image of the small-scale structure of soil, exhibiting polydispersity and anisotropy that result from a combination of erosive, compaction, and transport actions.

@@

More generally, the coupling between shape and flow during erosion, dissolution, and related phase-change processes is a topic that has garnered recent interest in the mathematical sciences community~\cite{Ristroph2012, rycroft2016asymmetric, cohen2016erosion, MooreCPAM2017, mitchell2017generalized, Quaife2018, wykes2018self, lachaussee2018competitive, lopez2018cfd, ristroph2018sculpting}.  Erosion-induced shape change has been studied in the high Reynolds number regime, both experimentally~\cite{Ristroph2012, MoorePOF2013} and theoretically~\cite{MoorePOF2013, MooreCPAM2017, hewett2017evolution} and in single~\cite{Ristroph2012, MoorePOF2013, MooreCPAM2017, lachaussee2018competitive} and multiple-body arrangements~\cite{hewett2017evolution}.  Studies on related processes of dissolution~\cite{HuangJFM2015, kondratiuk2015steadily, cohen2016erosion, claudin2017dissolution, wykes2018self, huang2017solute}, melting~\cite{Cummings1999, back2014effect, rycroft2016asymmetric, hewett2017pear, MooreCPAM2017, toppaladoddi2018combined, hewett2019evolution, favier2019rayleigh}, solidification~\cite{Huppert90, Anderson95, Cummings1999, mccue2003extinction}, and deposition~\cite{hewett2018modelling} exhibit intriguing similarities and differences~\cite{MooreCPAM2017}.  These studies reveal a rich set of possible dynamics, from the emergence of universal morphologies that erase details of the initial state and evolve self-similarly in time~\cite{MoorePOF2013, MooreCPAM2017, hewett2017pear}, to fine-scale pattern formation~\cite{cohen2016erosion, claudin2017dissolution, huang2017solute}, to unstable morphologies that can retain or even amplify features present in the initial state~\cite{claudin2017dissolution, wykes2018self}.

@@

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies with a volume fraction of $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities, for example contaminants.

@@

In addition to the work of the PIs, there has been a recently sparked interest in dynamic erosion of porous media, as experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing occur over much shorter timescales.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models~\cite{Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.

%----------------------------------------------------------------------%
\subsection{Governing Equations}

Here, we first outline the theoretical foundation to be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations \begin{equation}
\label{eqn:StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the
fluid viscosity, and $\Omega$ is the fluid domain. Linear elliptic PDEs,
such as the Stokes equations, can be reformulated as a boundary integral
equation (BIE).  BIEs have several numerical advantages including a
dimension reduction, automatic mass conservation, and high-order
accuracy. An equivalent BIE formulation of the Stokes equations in
Eq.~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Eq.~\eqref{eqn:StokesEq}, and its equivalent integral equation formulation, are microscopic descriptions that are required to study the shape evolution of individual grains due to fine-scale erosive effects, either mechanical or chemical. The description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

While the most immediate proposed work on computational methods to
simulate erosion will utilize~\eqref{eqn:StokesEq}, the later stages of this project will also incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models~\cite{Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed work of porous-media erosion. Large ensembles of the high-fidelity simulations will inform the coarse-grained models that will be developed in the later stages of the proposed work.
  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Scientific Computing, Mathematics, and others.
  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close cooperation between them.
\end{itemize}

@@

\paragraph{PIs' Qualifications:} PI Moore joined the Department of Mathematics at Florida State University (FSU) in 2014. He is also a Research Associate in FSU's Geophysical Fluid Dynamics Institute (GFDI).  His expertise includes modeling dynamic interactions between fluids and structures, especially in geophysical and biologically motivated settings. PI Quaife joined the Department of Scientific Computing at FSU in 2015 and is also a Research Associate in GFDI. His expertise includes high-fidelity and efficient numerical methods for fluid dynamic in complex geometries.

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, which is the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion \cite{chi-moo-qua2019}.

@@

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables closer contact between bodies and thus more bodies. The color in Fig~\ref{80circ8vort} shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity (Fig.~\ref{80circ8vel}).  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows new, preliminary results on how media properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies erode. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more mild decrease in area fraction is not sufficient to account for this exponential rate, and hence the reshaping process, in particular the formation of channels, plays a pivotal role.

Lastly, the right-most plot of Fig.~\ref{fig3} shows the anisotropy, which is calculated as the ratio of the vertical resistivity to the horizontal resistivity. The vertical resistivity is computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

@@

{An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations.} A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. However, the implementation of the quadrature method does not fully use FMM acceleration. The numerical challenge is developing the required FMM expansions for the velocity, vorticity, and deformation tensor layer potentials. Since we require several layer potentials with different kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and its HPC implementations that are readily available (for example, see \url{http://pvfmm.org}). Other challenges that will be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to allow for stable simulations to user-specified tolerances. PI Quaife has experience developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.

>>>>

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, i.e.~dissolution, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Some examples in the figure include the rough surface of limestone pavement, a sinkhole relic in Belize, and an erosion-curved conduit. The final image shows the small-scale structure of soil which exhibits polydispersity and anisotropy resulting from a combination of erosive, compaction, and transport actions.

@@

More generally, the coupling between shape and flow during erosion, dissolution, and related phase-change processes is a vibrant research area in the mathematical sciences community~\cite{Ristroph2012, rycroft2016asymmetric, cohen2016erosion, MooreCPAM2017, mitchell2017generalized, Quaife2018, wykes2018self, lachaussee2018competitive, lopez2018cfd, ristroph2018sculpting}.  Erosion-induced shape change has been studied in the high Reynolds number regime, both experimentally~\cite{Ristroph2012, MoorePOF2013} and theoretically~\cite{MoorePOF2013, MooreCPAM2017, hewett2017evolution} and in single~\cite{Ristroph2012, MoorePOF2013, MooreCPAM2017, lachaussee2018competitive} and multiple-body arrangements~\cite{hewett2017evolution}.  Studies on related processes of dissolution~\cite{HuangJFM2015, kondratiuk2015steadily, cohen2016erosion, claudin2017dissolution, wykes2018self, huang2017solute}, melting~\cite{Cummings1999, back2014effect, rycroft2016asymmetric, hewett2017pear, MooreCPAM2017, toppaladoddi2018combined, hewett2019evolution, favier2019rayleigh}, solidification~\cite{Huppert90, Anderson95, Cummings1999, mccue2003extinction}, and deposition~\cite{hewett2018modelling} exhibit intriguing similarities and differences~\cite{MooreCPAM2017}.  These studies reveal a rich set of possible dynamics, from the emergence of universal morphologies that erase details of the initial state and evolve self-similarly in time~\cite{MoorePOF2013, MooreCPAM2017, hewett2017pear}, to fine-scale pattern formation~\cite{cohen2016erosion, claudin2017dissolution, huang2017solute}, to unstable morphologies that can retain or even amplify features present in the initial state~\cite{claudin2017dissolution, wykes2018self}.

@@

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

@@

In addition to the work of the PIs, dynamic erosion of a porous media is a highly active subject as laboratory experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing represent rapid changes.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.

%----------------------------------------------------------------------%
\subsection{Theoretical Foundation and Governing Equations}

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations \begin{equation}
\label{eqn:StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the
fluid viscosity, and $\Omega$ is the fluid domain. Linear elliptic PDEs,
such as the Stokes equations, can be reformulated as a boundary integral
equation (BIE).  BIEs have several numerical advantages including a
dimension reduction, automatic mass conservation, and high-order
accuracy. An equivalent BIE formulation of the Stokes equations in
Eq.~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

Equation \eqref{eqn:StokesEq}, and its equivalent integral formulation, are microscopic descriptions required to model the shape evolution of individual grains due to fine-scale erosive effects, either mechanical or chemical. This microscopic description, however, becomes impractical when attempting to model seepage in very large porous systems where the microscopic geometry can be exceedingly complex. Instead, one must homogenize over microscopic details to arrive at an approximate description. Leading-order homogenization produces the well-known Darcy system 
\begin{align}
  \label{DarcyEq}
  \qq = - \bvec{K} \grad \phi
\end{align}
where $\qq$ is the {\em specific discharge} or {\em Darcy velocity}, and $\phi = z + p/(\rho g)$ is the hydraulic head, with $\rho$ and $g$ representing the fluid density and gravitational constant respectively.  Above, $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ through $\bvec{K} = \bvec{k} \rho g/\mu$.  Both the conductivity and permeability are in general tensors, and they reduce to scalars in the case of an isotropic medium. Note that the Darcy velocity is related to the tracer velocity through $\qq = \theta_f \uu$, where $\theta_f$ is porosity (i.e.~fluid volume fraction).  In the case of an isotropic medium, the Kozeny-Carman equation is often used to relate permeability to volume fraction~\cite{bear2013dynamics}
% See Bear Ch. 2.8, pg. 111
\begin{align}
  k = C \frac{\theta_f^3}{(1-\theta_f)^2}.
\end{align}
The macroscopic description offered by Eq.~\eqref{DarcyEq}, although approximate, simplifies analysis tremendously in that individual grains are no longer resolved and thus the domain geometry is generally much simpler.

Whilst the most immediate proposed work on computational methods to
simulate erosion will utilize~\eqref{eqn:StokesEq}, the later stages of this project will also incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed research on porous-media erosion and dispersive transport. Large ensembles of the high-fidelity simulations will inform the coarse-grained models to be developed in the later stages of the proposed work.
  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Scientific Computing, Mathematics, and others.
  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close interaction and cooperation between them.
\end{itemize}

@@

\paragraph{PIs' Qualifications:} PI Moore joined the Department of Mathematics at Florida State University (FSU) in 2014. He is also a Research Associate in FSU's Geophysical Fluid Dynamics Institute (GFDI).  His expertise includes modeling dynamic interactions between fluids and structures, especially in geophysical and biologically motivated settings. PI Quaife joined the Department of Scientific Computing at FSU in 2015 and is also a Research Associate in GFDI. His expertise includes high-fidelity and efficient numerical methods for fluid dynamics in complex geometries.

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion \cite{chi-moo-qua2019}.

@@

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables more densely packed configurations. The color in Fig~\ref{80circ8vort} represents the vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity.  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. As erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows how certain medium properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies disintegrate. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more modest decrease of area fraction is not sufficient to account for this exponential rate. Hence the reshaping process, in particular the formation of channels, plays a pivotal role the medium's macroscopic resistivity to flow.

Lastly, the right-most plot of Fig.~\ref{fig3} shows new, preliminary results on how medium anisotropy evolves during the reshaping process of erosion. Anisotropy calculated as the ratio of vertical to horizontal resistivity, with the vertical resistivity computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

@@

{An immediate goal of the proposed work is to extend the range of physical effects captured in these simulations.} A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, permitting an error bound that is independent of the grain configuration. However, the current implementation does not yet fully exploit FMM acceleration. The numerical challenge here is to develop FMM expansions for the velocity, vorticity, and deformation-tensor layer potentials. Since these layer potentials each have distinct kernels, we will utilize kernel-independent FMMs~\cite{yin-bir-zor2004, fon-dar2009}, and their HPC implementations (for example, see \url{http://pvfmm.org}). Other challenges to be investigated include preconditioning strategies to reduce the number of required GMRES iterations, and time adaptivity to enable stable simulations to user-specified tolerances. PI Quaife is a leading expert in developing such algorithms~\cite{qua-bir2015b, qua-bir2016, qua-bir2015a, qua-cou-dar2018}.

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations and the {\thL} method for interface evolution, is already in place. The BIE Stokes solver, however, has not yet been coupled to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} required for dissolution.  We proposed to solve the advection-diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the system into individual PDEs for linear advection and diffusion.  The advection equation will be handled with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience using semi-Lagrangian methods in the closely related context of complex viscous flow~\cite{kab-qua-bir2017}.

The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0.
\end{alignat}
The transformed variable $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega, \: s \in \mathbb{C},
  \label{eqn:DiffusionLaplace}
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}. The concentration field can then be recovered with an inverse Laplace transform~\cite{jos-war2006}.

By combining our BIE Stokes solver with the proposed high-order, advection-diffusion solver, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible.  Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes (see Section~\ref{sec:sinkhole}).
